A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-22 inst/manuscript/manuscript.Rmd
more minor updates

We used a simulation study to investigate performance of four sampling-analysis combinations: Independent Random Sampling (IRS) with a design-based analysis, called "IRS-Design" (IRS samples are not spatially balanced because population units are selected independent of their spatial locations); IRS with a model-based analysis, called "IRS-Model"; GRTS sampling with a design-based analysis, called "GRTS-Design"; and GRTS sampling with a model-based analysis, called "GRTS-Model". These combinations are also provided in Table \ref{tab:designanalysis}.

@@

Performance for the four sampling-analysis combinations was evaluated in 36 different simulation scenarios. The 36 scenarios resulted from the crossing of three sample sizes, two location layouts (of the population units), two response types, and three proportions of dependent random error. The three sample sizes ($n$) were $n = 50, n = 100,$ and $n = 200$. Samples were always selected from a population size ($N$) of $N = 900$. The two location layouts were random and gridded. Locations in the random layout were randomly generated inside the unit square ($[0, 1] \times [0, 1]$). Locations in the gridded layout were placed on a fixed, equally spaced grid inside the unit square. The two response types were normal and lognormal. For the normal response type, the response was simulated using mean-zero random errors with the exponential covariance (Equation$~$\ref{equation:expcov}) for varying proportions of dependent random error. The proportion of dependent random error is represented by $\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)$, where $\sigma^2_1$ and $\sigma^2_2$ are the dependent random error variance (partial sill) and independent random error variance (nugget), respectively, from Equation$~$\ref{equation:expcov}. The total variance, $\sigma^2_1 + \sigma^2_2$, was always 2. The range was always $\sqrt{2} / 3$, which means that the correlation in the dependent random error decayed to nearly zero at the largest possible distance between two population units in the domain. For the lognormal response type, the response was first simulated using the same approach as for the normal response type, except that the total variance was 0.6931 instead of 2. The response was then exponentiated, yielding a lognormal random variable whose total variance is 2. The lognormal responses were used to evaluate performance of the sampling-analysis approaches for data that were skewed (i.e., not normal).

@@

In each of the 36 simulation scenarios, there were 2000 independent simulation trials. In each trial, IRS and GRTS samples were selected and then design-based and model-based analyses were used to estimate (design-based) or predict (model-based) the mean and construct 95% confidence (design-based) or 95% prediction (model-based) intervals. Then we recorded the bias, squared error, and interval coverage for all sampling-analysis combinations. After all 2000 trials, we summarized the long-run performance of the combinations by calculating average bias, rMS(P)E (root-mean-squared error for the design-based approaches and root-mean-squared-prediction error for the model-based approaches), and the proportion of times the true mean is contained in its 95% interval. The GRTS algorithm and the local neighborhood variance estimator are available in the `spsurvey` \textbf{\textsf{R}} package [@dumelle2021spsurvey]. FPBK is available in the `sptotal` \textbf{\textsf{R}} package [@higham2021sptotal] and covariance parameters were estimated using Restricted Maximum Likelihood  [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing].

@@

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the contiguous United States. We will use data from the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and water quality [@USEPA2012NLA]. Specifically, we will analyze mercury concentration in lakes. Although we can calculate the true mean mercury concentration values for the 986 lakes from the 2012 NLA, we will explore whether or not we obtain an adequately precise estimate for the realized mean mercury concentration if we sample only 100 of the 986 lakes. For each of the four familiar sampling-analysis combinations (IRS-Design, IRS-Model, GRTS-Design, and GRTS-Model), we estimate (design-based) or predict (model-based) the mean mercury concentration and construct 95% confidence (design-based) or 95% prediction (model-based) intervals from this sample of 100 lakes, which we compare to the true mean mercury concentration from all 986 lakes.

@@

Fig. \ref{fig:figeff} shows the relative rMS(P)E of the four sampling analysis combinations using the random location layout with "IRS-Design" as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}
When there is no spatial covariance (Fig. \ref{fig:figeff}, "Prop DE: 0" row), the four sampling-analysis combinations have approximately equal rMS(P)E. So using the GRTS sampling plan or a model-based analysis does not result in much, if any, loss in efficiency compared to IRS-Design when there is no spatial covariance. When there is spatial covariance (Fig. \ref{fig:figeff}, "Prop DE: 0.5" and "Prop DE: 0.9" rows), GRTS-Model tends to perform best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and IRS-Model is relatively small.  As the strength of spatial covariance increases, the gap in rMS(P)E between IRS-Design and the other sampling-analysis combinations widens. Finally we note that when there is spatial covariance, IRS-Model outperforms IRS-Design by a large margin, suggesting that the poor design properties of IRS are largely mitigated by the model-based analysis. These conclusions are similar to those observed in the grid location layout, so we omit a grid location layout figure here. Tables for rMS(P)E in all 36 simulation scenarios are provided in the supporting information.
\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative rMS(P)E in the simulation study for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}
  \label{fig:figeff}
\end{figure}

@@

We selected a single IRS sample and a single GRTS sample and estimated (design-based) or predicted (model-based) the mean mercury concentration and constructed 95% confidence (design-based) and 95% (model-based) prediction intervals. For the model-based analyses, the exponential covariance was used. Table \ref{tab:appliedtab} shows the results from these analyses. For all four sampling-analysis combinations, the true realized mean mercury concentration is within the bounds of the 95% confidence (design-based) or 95% prediction (model-based) intervals. Though we should not generalize these results to other samples from these data, we do note a couple of patterns. The design-based IRS analysis shows the largest standard error: a likely reason is that this is the only approach that does not incorporate any spatial information regarding mercury concentration. Both analyses using GRTS sampling have lower standard errors than both analyses using IRS sampling. We expect that these patterns are consistent with other samples from these data because mercury concentration exhibits spatial dependence, so a spatially balanced sample should usually yield a lower standard error.

@@

The design-based and model-based approaches to statistical inference are fundamentally different paradigms. The design-based approach incorporates randomness through sampling to estimate population parameters. The model-based approach incorporates randomness through distributional assumptions to predict realized values of a stochastic process. Though these approaches have often been compared in the literature from theoretical and analytical perspectives, our contribution lies in studying them in a spatial context while implementing spatially balanced sampling and the local neighborhood variance estimator (in the design-based approach). Aside from the theoretical differences described, a few analytical findings from the simulation study are particularly notable. First, the sampling decision (IRS vs GRTS) is most important when using a design-based analysis. Though GRTS-Model still outperformed IRS-Model, the model-based analysis mitigated much of the inefficiency of the IRS sample. Second, independent of the analysis approach, we found no reason to prefer IRS over GRTS for sampling spatial data -- GRTS-Design and GRTS-Model generally performed at least as well as their IRS counterparts when there was no spatial covariance and noticeably better than their IRS counterparts when there was spatial covariance. Third, as the strength of spatial covariance increases, the gap in rMS(P)E between IRS-Design and the other sampling-analysis combinations also increases. Fourth and finally, when the response was normal, interval coverage for all sampling-analysis combinations was very close to 95% for all sample sizes; when the response was lognormal, interval coverage for all sampling and analysis was between 90% and 95% and closest to 95% when $n = 200$.

There are several benefits and drawbacks of the design-based and model-based approaches for finite population spatial data. Some we have discussed, but others we have not, and they are worthy of consideration in future research.  Design-based approaches are often computationally efficient, while model-based approaches can be computationally burdensome, especially for likelihood-based estimation methods like REML that rely on inverting a covariance matrix. The design-based approach also more naturally handles binary data, free from the more complicated logistic regression framework commonly used to analyze binary data in a model-based approach. The model-based approach, however, can more naturally quantify the relationship between covariates (predictor variables) and response variable. The model-based approach also yields estimated spatial covariance parameters, which help better understand the dependence structure in the stochastic process of study. Model selection is also possible using model-based approaches and criteria such as cross validation, likelihood ratio tests, or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed sites. Model-based approaches can also compute site-by-site predictions at unobserved locations and use them to construct informative visualizations. In short, when deciding whether the design-based or model-based approach is more appropriate to implement, the benefits and drawbacks of each approach should be considered alongside the particular goals of the study.

>>>>

We used a simulation study to investigate performance of four sampling-analysis combinations. The first sampling-analysis combination is IRS-Design. In IRS-Design, samples are selected using the Independent Random Sampling (IRS) algorithm. The IRS algorithm ignores the spatial locations of the population units, which implies IRS samples are not spatially balanced. In IRS-Design, samples are analyzed using the design-based approach with an IRS variance estimator that does not incorporate the spatial locations of the units in the sample. The second sampling-analysis combination is IRS-Model, where samples are selected using the IRS algorithm and analyzed using the model-based approach via Restricted Maximum Likelihood (REML) estimation [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing]. The third sampling-analysis combination is GRTS-Design, where samples are selected using the GRTS algorithm and analyzed using the design-based approach with the local neighborhood variance estimator. The fourth and final sampling-analysis combination is GRTS-Model, where samples are selected using the GRTS algorithm and analyzed using the model-based approach via REML estimation. These sampling-analysis combinations are also provided in Table \ref{tab:designanalysis}. Lastly we note that for both the IRS and GRTS samples, equal inclusion probabilities were assumed for all population units. When IRS assumes equal inclusion probabilities for all population units, the algorithm is equivalent to "simple random sampling."

@@

Performance for the four sampling-analysis combinations was evaluated in 36 different simulation scenarios. The 36 scenarios resulted from the crossing of three sample sizes, two location layouts (of the population units), two response types, and three proportions of dependent random error. The three sample sizes ($n$) were $n = 50, n = 100,$ and $n = 200$. Samples were always selected from a population size ($N$) of $N = 900$. The two location layouts were random and gridded. Locations in the random layout were randomly generated inside the unit square ($[0, 1] \times [0, 1]$). Locations in the gridded layout were placed on a fixed, equally spaced grid inside the unit square. The two response types were normal and lognormal. For the normal response type, the response was simulated using mean-zero random errors with the exponential covariance (Equation$~$\ref{equation:expcov}) for varying proportions of dependent random error. The proportion of dependent random error is represented by $\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)$, where $\sigma^2_1$ and $\sigma^2_2$ are the dependent random error variance (partial sill) and independent random error variance (nugget), respectively, from Equation$~$\ref{equation:expcov}. The total variance, $\sigma^2_1 + \sigma^2_2$, was always 2. The range was always $\sqrt{2} / 3$, which means that the correlation in the dependent random error decayed to nearly zero at the largest possible distance between two population units in the domain. For the lognormal response type, the response was first simulated using the same approach as for the normal response type, except that the total variance was 0.6931 instead of 2. The response was then exponentiated, yielding a lognormal random variable whose total variance was 2. The lognormal responses were used to evaluate performance of the sampling-analysis approaches for data that were skewed (i.e., not normal).

@@

In each of the 36 simulation scenarios, there were 2000 independent simulation trials. In each trial, IRS and GRTS samples were selected and then design-based and model-based analyses were used to estimate (design-based) or predict (model-based) the mean and construct 95% confidence (design-based) or 95% prediction (model-based) intervals. Then we recorded the bias, squared error, and interval coverage for all sampling-analysis combinations. After all 2000 trials, we summarized the long-run performance of the combinations by calculating average bias, rMS(P)E (root-mean-squared error for the design-based approaches and root-mean-squared-prediction error for the model-based approaches), and the proportion of times the true mean is contained in its 95% interval. The IRS algorithm, IRS variance estimator, GRTS algorithm, and local neighborhood variance estimator are available in the `spsurvey` \textbf{\textsf{R}} package [@dumelle2021spsurvey]. FPBK is available in the `sptotal` \textbf{\textsf{R}} package [@higham2021sptotal].

@@

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the contiguous United States. One component of NARS is the National Lakes Assessment (NLA), which measures various aspects of lake health and water quality [@USEPA2012NLA]. We will analyze mercury concentration data collected at 986 lakes as part of the 2012 NLA. Although we can calculate the true mean mercury concentration values for these 986 lakes, here we will explore whether or not we can obtain an adequately precise estimate for the realized mean mercury concentration if we sample only 100 of the 986 lakes. For each of the four familiar sampling-analysis combinations (IRS-Design, IRS-Model, GRTS-Design, and GRTS-Model), we estimate (design-based) or predict (model-based) the mean mercury concentration and construct 95% confidence (design-based) or 95% prediction (model-based) intervals from this sample of 100 lakes, which we compare to the true mean mercury concentration from all 986 lakes.

@@

Fig. \ref{fig:figeff} shows the relative rMS(P)E of the four sampling analysis combinations using the random location layout with "IRS-Design" as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}
When there is no spatial covariance (Fig. \ref{fig:figeff}, "Prop DE: 0" row), the four sampling-analysis combinations have approximately equal rMS(P)E. So using the GRTS algorithm or a model-based analysis does not result in much, if any, loss in efficiency compared to IRS-Design when there is no spatial covariance. When there is spatial covariance (Fig. \ref{fig:figeff}, "Prop DE: 0.5" and "Prop DE: 0.9" rows), GRTS-Model tends to perform best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and IRS-Model is relatively small.  As the strength of spatial covariance increases, the gap in rMS(P)E between IRS-Design and the other sampling-analysis combinations widens. Finally we note that when there is spatial covariance, IRS-Model outperforms IRS-Design by a large margin, suggesting that the poor design properties of IRS are largely mitigated by the model-based analysis. These conclusions are similar to those observed in the grid location layout, so we omit a grid location layout figure here. Tables for rMS(P)E in all 36 simulation scenarios are provided in the supporting information.
\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative rMS(P)E in the simulation study for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}
  \label{fig:figeff}
\end{figure}

@@

We selected a single IRS sample and a single GRTS sample and estimated (design-based) or predicted (model-based) the mean mercury concentration and constructed 95% confidence (design-based) and 95% (model-based) prediction intervals. For the model-based analyses, the exponential covariance was used. Table \ref{tab:appliedtab} shows the results from these analyses. For all four sampling-analysis combinations, the true realized mean mercury concentration is within the bounds of the 95% confidence (design-based) or 95% prediction (model-based) intervals. Though we should not generalize these results to other samples from these data, we do note a couple of patterns. The design-based IRS analysis shows the largest standard error: a likely reason is that this is the only approach that does not incorporate any spatial locations. Additionally, both analyses using GRTS sampling have lower standard errors than both analyses using IRS sampling.

@@

The design-based and model-based approaches to statistical inference are fundamentally different paradigms. The design-based approach incorporates randomness through sampling to estimate population parameters. The model-based approach incorporates randomness through distributional assumptions to predict realized values of a stochastic process. Though these approaches have often been compared in the literature from theoretical and analytical perspectives, our contribution lies in studying them in a spatial context while implementing spatially balanced sampling and the local neighborhood variance estimator (in the design-based approach). Aside from the theoretical differences described, a few analytical findings from the simulation study are particularly notable. First, the sampling decision (IRS vs GRTS) is most important when using a design-based analysis. Though GRTS-Model still outperformed IRS-Model, the model-based analysis mitigated most of the inefficiencies that result from the IRS samples lacking spatial balance. Second, independent of the analysis approach, we found no reason to prefer IRS over GRTS for sampling spatial data -- GRTS-Design and GRTS-Model generally performed at least as well as their IRS counterparts when there was no spatial covariance and noticeably better than their IRS counterparts when there was spatial covariance. Third, as the strength of spatial covariance increases, the gap in rMS(P)E between IRS-Design and the other sampling-analysis combinations also increases. Fourth and finally, when the response was normal, interval coverage for all sampling-analysis combinations was very close to 95% for all sample sizes; when the response was lognormal, interval coverage for all sampling and analysis was between 90% and 95% and closest to 95% when $n = 200$.

There are several benefits and drawbacks of the design-based and model-based approaches for finite population spatial data. Some we have discussed, but others we have not, and they are worthy of consideration in future research.  Design-based approaches are often computationally efficient, while model-based approaches can be computationally burdensome, especially for likelihood-based estimation methods like REML that rely on inverting a covariance matrix. The design-based approach also more naturally handles binary data, free from the more complicated logistic regression framework commonly used to analyze binary data in a model-based approach. The model-based approach, however, can more naturally quantify the relationship between covariates (predictor variables) and response variable. The model-based approach also yields estimated spatial covariance parameters, which help better understand the dependence structure in the stochastic process of study. Model selection is also possible using model-based approaches and criteria such as cross validation, likelihood ratio tests, or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed sites. Model-based approaches can also compute site-by-site predictions at unobserved locations and use them to construct informative visualizations like smoothed maps. In short, when deciding whether the design-based or model-based approach is more appropriate to implement, the benefits and drawbacks of each approach should be considered alongside the particular goals of the study.
